Modeling electrostatics in molecular biology: A tutorial of DelPhi and associated resources [Article v\versionnumber]

### 2019-09-16 article-manuscript/article-manuscript-tutorial-v01.tex
English corrections

Electrostatics play indispensable role in practically any process in molecular biology. Indeed, at distances larger than several Angstroms all other forces are negligibly small and electrostatic force dominates. However, modeling electrostatics in molecular biology is a complicated task due to presence of water phase, mobile ions and irregularly shaped inhomogeneous biological macromolecules. A particular approach to calculating electrostatics in such systems is to apply Poisson-Boltzmann equation (PBE). Here we provide a tutorial for the popular DelPhi package that solves PBE using finite-difference method and delivers the electrostatic potential distribution throughout the modeling box. The tutorial comes with detailed description of different tasks that DelPhi can handle, an assessment of the accuracy against cases with analytical solution and recommendations about DelPhi usage. Furthermore, since electrostatics is a key component of virtually any modeling in molecular biology, we have created many additional resources utilizing DelPhi to model various biology relevant quantities. Tutorials for these resources are also provided along with examples of their usage. 
\end{abstract}

@@

\section{Introduction}
Poisson-Boltzmann equation (PBE) is widely used to describe the electrostatic potential distribution in systems made of biological macromolecules in water phase in presence of mobile ions\cite{xiao2017continuum,li2013progress,jurrus2018improvements,nguyen2017accurate}. Since in this approach the macromolecules are considered using their atomistic 3D structures, it represents an excellent trade off between the speed and the necessary details for accurate description of the relevant phenomena. The traditional implementation of PBE in molecular biology considers that biological macromolecules are low dielectric cavities immersed in water phase which is described as continuum medium with large dielectric constant. Ions are modeled in such a protocol as point charges that obey Boltzmann distribution and can be present outside of so termed Stern Layer (Figure \ref{fig:Figure_1}a). Such an implementation of PBE will be termed traditional PBE. In parallel, we have implemented Gaussian-based smooth dielectric function in DelPhi that treats the solute and solvent on the same footage and the entire computational space is described via continuous dielectric function\cite{chakravorty2018gaussian,jia2017treating,li2014modeling,li2013dielectric}. Thus, the macromolecules are considered to be inhomogeneous objects and there is no sharp boundary between solute-solvent. Ions in such a scenario are considered as point charges obeying Boltzmann distribution, but the argument of the Boltzmann function has a desolvation penalty which does not allow ions to propagate into macromolecular interior unless there is a cavity (Figure \ref{fig:Figure_1}b). This approach will be termed Gaussian PBE. Both approaches have been used in various computational algorithms aimed at computing biologically important quantities and these resources will be described here as well along with appropriate tutorials.

The DelPhi software and associated resources are available free of charge for either download as a stand-alone code or used as webservers. The general access URL is \url{http://compbio.clemson.edu}.

@@

\subsubsection{Continuum electrostatics quantities} The key electrostatic quantity discussed in this tutorial is electrostatic potential distribution throughout the space. The space is occupied by macromolecule(s) and water phase. In terms of continuum electrostatics, the macromolecules and water phase are considered as continuum media. For such a system, the electrostatic potential obeys various differential equations, which will be described below. Essential term in these equations is so termed dielectric “constant” or dielectric function. In some models the dielectric function is a constant in macromolecular and water media, while in other models it is a smooth continuous function. The macromolecules are considered at atomistic level of details, and atomistic radii and charges are assigned according appropriate force field parameters. Finally, the mobile ions present in the water phase are considered point charges and their distribution is governed by the Boltzmann law. Below we outline two major models of partitioning the computational space.

\subsubsection{Space partitioning in case of standard two-dielectric model}. The space partitioning\cite{rocchia2002rapid,baker2001electrostatics} and the corresponding differential equations being solved in case of traditional PBE involve splitting the space into three regions labeled as $ \Omega_1 $, $ \Omega_2 $ and $ \Omega_3 $ in Figure \ref{fig:Figure_1}a. The first space region, the $\Omega_1$ is occupied by the macromolecule(s) and its boundary is defined as the molecular surface. It is considered to be a low dielectric cavity with dielectric constant of $\epsilon_1$. Molecular surface itself may be constructed via various approaches, which are not subject of this manuscript. The outer space region, the $\Omega_3$ region, is taken by the bulk water, considered to be continuum medium with dielectric constant of $\epsilon_3$. In-between is space region $\Omega_2$, which is termed Stern layer. This region has the same dielectric properties as the bulk water, but ions are not allowed to penetrate there (Figure \ref{fig:Figure_1}a).

\begin{figure*}[hbt!]
\includegraphics[width=0.95\linewidth]{Figure_1.png}
\caption{\textbf{(a)} Schematic presentation of space partitioning in case of traditional 2-dielectric model. In each of these partitions, different differential equation is solved. The domain, which is essentially the 3D space containing the macromolecule(s) and the solvent (with or without ions) is split into three regions labelled as $ \Omega_1$, $ \Omega_2 $ and $ \Omega_3 $. \textbf{(b)} Schematic presentation of the dielectric distribution delivered using the Gaussian-based smooth dielectric model (figure shows dielectric function value throughout a slice of a protein). Note that there is no dielectric boundary and even some space regions inside the protein are assigned high dielectric constant.}
\label{fig:Figure_1}
\end{figure*}

@@

The outer space region or the $ \Omega_3 $ is occupied by the bulk of the solvent (typically water) which represents a continuum medium with a dielectric value $ \epsilon_3 $. The solvent continuum can also contain mobile ions whose effects are implicitly taken into account using the Boltzmann law. The potential distribution in this regions is governed by the Poisson-Boltzmann equation (PBE). Therefore, here, the charge density, $ \rho_3(\Vec{r}) $, due to the mobile ions (with concentration M) generates a potential distribution that can be expressed as Eqn. \ref{eqn:pbe_1}.

@@

In the above equations, $ e_c $ is the elementary charge, $ k_B $ is the Boltzmann constant and $ T $ is the absolute temperature.
In between these two regions exists the ion-exclusion layer (shown as $ \Omega_2 $) which the mobile ions are not allowed to access. Due to the lack of mobile ions as well as the absence of permanent solute charges, this region is governed by Laplace’s equation.
\begin{equation}
\nabla^2\phi(\Vec{r}) = 0
\end{equation}
The ion-exclusion layer or the Stern layer has the same dielectric properties as the solvent region or $ \Omega_3 $ (i.e. $ \epsilon_3 $). Therefore, $ \epsilon_2=\epsilon_3 $ . This indicates the role of the molecular surface as the surface that separates two dielectric media. The molecular surface can be constructed using various approaches but they are out of the scope of this article\cite{rocchia2002rapid,chakravorty2019grid,decherchi2013between}.

\subsubsection{Space partitioning with the Gaussian-based smooth dielectric model}
The Gaussian-based smooth dielectric model is designed to deliver a continuous distribution of dielectric values in space as opposed to representing it as a union of two media with distinct dielectric constants. 
The smooth distribution reflects the effect of restrictions on the mobility of the macromolecular and solvent regions on their polarizability or the dielectric value. The distribution is assigned based on the macromolecular atomic packing density in a given region. Its formulation ensures that the dielectric value is the lowest ($ \epsilon_1 $) at the center of the macromolecular atoms and is the highest ( $ \epsilon_2 $ ) in the bulk. Through this assignment, regions close to the macromolecular atom centers attain a lower dielectric value and those farther away from the macromolecule attain a higher value. Overall, buried or highly densely packed regions of the system feature a lower dielectric value than those which are less packed (details are provided in refs\cite{chakravorty2019grid,decherchi2013between}). This is illustrated in Figure \ref{fig:Figure_1}b.

The key features of Gaussian-based smooth dielectric functions are that the strict surface (like the molecular surface shown in Figure \ref{fig:Figure_1}a) is replaced by a smooth transition of the dielectric values from the macromolecular interior $ \Omega_1 $ to the bulk $ \Omega_3 $ as shown in Figure \ref{fig:Figure_1}). The transient values of the dielectric at the solute-solvent interface collectively reflects the increased polarizability of solvent exposed atoms of the macromolecule (compared to those in the core regions) and decreased polarizability of the solvent in their vicinity (compared to those in the bulk). This occurs by virtue of the solute-solvent interactions that play a critical role in balancing the effects of (de)solvation and intramolecular interactions of the solvent exposed macromolecular atoms. 
Due to the continuous dielectric distribution, a clear distinction of macromolecular and solvent regions is no longer necessary and therefore, the same differential equation (PE or PBE) can be used homogeneously. However, the lack of a dielectric boundary invites a modification in the way by which the screening effect of mobile ions is taken into account. In the traditional 2-dielectric approach, the presence of a mobile ion at a point in space (depending on its accessibility) is guided by the electrostatic potential at that point which is expressed using a Boltzmann distribution function (Eqn. \ref{eqn:pbe_2}). In the Gaussian-based approach, an additional penalty term is added to its potential energy at any point in space (for more details see Ref\cite{jia2017treating}). The penalty term is based on the Born-formalism of the energy of transfer of a centrosymmetric ion across two dielectric media. In the case of the Gaussian-model, the penalty term, $ \Delta G_{penalty} $, is the energy of transfer of an ion of valence $ z $ and Van der Waals radius $ R $ from the solvent (with dielectric $ \epsilon(\Vec{r})$) to a point  (with dielectric ) and is expressed in SI units as:

@@

In Figure \ref{fig:plot_saltation_penalty}, the penalty term as a function of position is shown for a solute with a very simple geometry (one sphere of 2 \text{\AA} radius). As one approaches the vicinity of the solute, the penalty increases dramatically, ensuring that the mobile ions cannot penetrate in the regions dominated by solute’s dielectric values. Farther out in the bulk, the penalty term diminishes to zero, thus retrieving the properties of the 2-dielecric approach.

@@

\subsubsection{Finite difference algorithm}
The Poisson-Boltzmann equation is a non-linear $2^{nd}$ order differential equation that is typically solved using numerical schemes since analytical solutions are only possible for cases with very simple geometry. Inhomogeneity of charge distribution and complex geometrical features are prerogative of biomolecules, therefore, different methods of numerically solving the PBE for biomolecular systems have been developed \cite{klapper1986focusing,davis1989solving,cortis1997numerical,baker2001electrostatics,shestakov2002solution,miertuvs1981electrostatic,totrov2001rapid,lu2009adaptive}. 
DelPhi solves the PBE using the finite difference (FD) algorithm. The approach rests on discretization of the entire space, $ \Omega=(\Omega_1 \cup \Omega_2 \cup \Omega_3)$, into a set of adjacent cubes, all of the same side length. Along each of the three Cartesian orthogonal directions, identical number of these cubes are placed and their union forms a larger cube which is also referred to as the computational box. Points where two or more cubes meet are known as the grid points and they are the locations where the electrostatic potentials and electric fields are determined.  Between two adjacently placed grid-points, a mid-point is located where the dielectric values are assigned. Collectively, the grid-points and mid-points are all placed within the bounds of the computational box. Figure \ref{fig:grid-scheme-representation}, provides an illustration of this discretization and a computational box with ‘$ M $’ cubes per side, each with a side length of ‘$ h $’.

@@

Instead of solving different variants of the differential equation like the PE and PBE (as mentioned above) in the different partitions or subdomains of the computational box, an iterative protocol based of the integral form of these equations is executed to compute the potential values at each and every grid-point. To do so, a position dependent formulation of the dielectric values ($ \epsilon $) and the modified Debye-Hückel parameter ($ \kappa $) is used.

@@

\subsection{Energy partitioning}
Once the potential distribution is calculated, the next step in DelPhi algorithm is to compute the electrostatic energies. Based on the original paper of Sharp and Honig\cite{sharp1990electrostatic}, see also\cite{rocchia2001extending}, the energy components are:
Grid energy
\begin{equation}
    G_{grid} = 0.5 \left[ \sum_i^{grids}{p_i\phi_i} \right]
\end{equation}
where $ p_i $ is the charge of grid point ‘$ i $’ and $ \phi_i $ is the grid potential at the same grid point. Note that this energy term contains artificial component due to the real charge partitioning over the grid points. Direct usage of $ G_{grid} $ is not appropriate, only grid energy difference can be used.

The total electrostatic energy of the system can be decomposed into three energy terms\cite{rocchia2001extending}, coulombic, polar solvation and saltation energies. In case of traditional two-dielectric model, this is described in Ref. \cite{rocchia2001extending}. Thus, the coulombic term is the product between real charges over the distance, and it is computed via Coulombic formula:

\begin{equation}
    G_{Coul} = 0.5 \left[ \sum_i^{\text{number of charges}}{\frac{q_i q_j}{\epsilon_{in}r_{ij}}} \right]
\end{equation}	
where $ q_i $ are the real charges, $r_{ij} $ is the distance between charge ‘$i$’ and ‘$j$’ and $ \epsilon_{in} $ is the internal dielectric constant.

The second energy term is the polar solvation energy, which in DelPhi is called “corrected reaction field energy, or $ G_{rxn} $”. Following original development due to Anthony Nicholls, this energy is calculated via induced charges on the molecular surface. Thus,

@@

The third term represents the energy of interactions between permanent charges and mobile ions. Formally it could be described by formula similar to Eqn. \ref{eqn:G_rxn}, where instead of induced charges one takes excess ion density. However, this approach was abandoned in the newest DelPhi version because it is very slow and requires small “percent fill”. Instead, we recommend using grid energy difference between runs with salt and without salt.

Finally, if one invokes non-linear PBE, two additional energy terms should be added to the total electrostatic energy. Following the work of Sharp and Honig\cite{sharp1990electrostatic}, they are termed “osmotic pressure” and electrostatic "stress”.

@@

DelPhi uses FD algorithm to solve the PBE and thus to deliver the electrostatic potential distribution. The first step in the algorithm requires that one provides a coordinate file of the biomolecules and nano-objects in PDB or PQR formats. In the second case, the input PQR files provides the coordinates of all atoms along with their radii and charges. In the first case, the user needs to provide two additional files, the “size” and “charge” files. These files contain information about the atomic radii and charges. If PDB format file is provided, then DelPhi matches the names of the atoms/residues toward the names/radii in the corresponding “size” and “charge” files and thus generates information for each atom position, radius and charge (note that if the PDB file contains atomic names/residues not available in the corresponding “size” and “charge” files, then DelPhi will generate warning message prompting the users that these names/residues are not recognized). Once the atomic coordinates and radii are available DelPhi builds grid presentation of the geometrical and dielectric properties of the system. By default, it is done by positioning the macromolecule(s) geometrical center at the center of the computational box and using “scale”, “grid size” and “percent fill”.

\begin{figure}[hbt!]
\includegraphics[width=\linewidth]{Figure_4.png}
\caption{The schematic representation of relationship between "percent fill" (\ttvar{perfil}), "grid size" (\ttvar{gsize}), and \ttvar{scale}. The parameter \ttvar{gsize} decides number of grid lines in each direction e.g. \ttvar{gsize} is 9 in given cartoon. The parameter \ttvar{scale} decides number of grid lines per \text{\AA}. If we assume perfil $ pf\% $ and maximum dimension in X-, Y- directions be $ L_{xp} $ \text{\AA} and $ L_{yp} $ \text{\AA} as shown in cartoon, then the size of the grid box $ L_x $ in \text{\AA} is computed using relation $ L_x = \frac{max(L_{xp}, L_{yp}) \times 100}{pf} $. The value of \ttvar{scale} in this case would be $ \ttvar{scale} = \frac{L_x}{\ttvar{gsize}} = \frac{L_x}{9} $.}
\label{fig:perfil-gsize-scale}
\end{figure}

Note that these three parameters, “scale”, “grid size” and “percent fill” are interdependent (Figure \ref{fig:perfil-gsize-scale}). Only two are needed. Users are advised to use “scale” and “percent fill” (\ttvar{perfil}) and to let DelPhi decide the “grid size” (\ttvar{gsize}). In the traditional two-dielectric model, the computational box is split into low dielectric cavities occupied by macromolecule(s) and water phase and appropriate dielectric constant values are assigned to the mid-point belonging to the corresponding phases. In the Gaussian model, coordinates and radii are used to construct a smooth Gaussian-based dielectric function for the entire computational box without distinguishing the macromolecule(s) from the water phase (for mode details see Refs \cite{li2014modeling,li2013dielectric}). Once this is completed, DelPhi initiates FD iterations to compute electrostatic potential distribution and to calculate the corresponding energy, as it will be outlined below. Several practical advice about DelPhi parameters are given here:

\begin{enumerate}
    \item How to decide on geometry? One wants the macromolecule(s) to occupy significant percent of the computational box space so the calculations to focus on important part of the system. However, doing modeling with very high “percent fill” may result in positioning some of the macromolecule(s) atoms too close to the edges of the computational box and thus to introduce an error. We recommend “percent fill” of about 70\%.

    \item How to decide on the "dielectric constants"? Most of the modeling involves biological macromolecules in water, so natural choice of the dielectric constant of solvent is the dielectric constant of bulk water, typically takes as 80. Regarding internal dielectric constant: (a) in case of traditional two-dielectric model, one selects the value of internal dielectric constant depending on the modeling protocol. If one models rigid structures or snap shots taken from MD simulations, the internal dielectric constant value is typically set to 1.0. However, in other modelings one may select values that are higher with the goal to match experimental data or to mimic for conformational flexibility not accounted in the protocol (b) in case of smooth Gaussian-based dielectric function, one selects the reference value of internal dielectric constant based on bench marking (see refs\cite{li2014modeling,li2013progress,wang2015pka,wang2015delphipka} for more details).

    \item Criteria for convergence. The FD algorithm is iterative algorithm and its convergence is monitored via several parameters as “maxc”, which is the largest change of the potential from iteration ‘i’ to ‘i+1’; or “rmsc”, which root mean square change of the potential. In both cases, it is recommended to use small, but reasonable criteria of 0.0001.

@@

Now close your terminal and restart terminal. Now user shall be able to perform calculations discussed in all of the examples compiled in directory \ttvar{LiveCoMS_Examples/}. All the paths mentioned in individual sections are relative to this directory.

>>>>

Electrostatics play an indispensable role in practically any process in molecular biology. Indeed, at distances larger than several Angstroms, all other forces are negligibly small and electrostatic force dominates. However, modeling electrostatics in molecular biology is a complicated task due to presence of water phase, mobile ions and irregularly shaped inhomogeneous biological macromolecules. A particular approach to calculating electrostatics in such systems is to apply the Poisson-Boltzmann equation (PBE). Here, we provide a tutorial for the popular DelPhi package that solves PBE using a finite-difference method and delivers the electrostatic potential distribution throughout the modeling box. The tutorial comes with a detailed description of different tasks that DelPhi can handle, an assessment of the accuracy against cases with analytical solutions and recommendations about DelPhi usage. Furthermore, since electrostatics is a key component of virtually any modeling in molecular biology, we have created many additional resources utilizing DelPhi to model various biology relevant quantities. Tutorials for these resources are also provided along with examples of their usage. 
\end{abstract}

@@

\section{Introduction}
Poisson-Boltzmann equation (PBE) is widely used to describe the electrostatic potential distribution in the systems made of biological macromolecules in the water phase in the presence of mobile ions\cite{xiao2017continuum,li2013progress,jurrus2018improvements,nguyen2017accurate}. Since in this approach the macromolecules are considered using their atomistic 3D structures, it represents an excellent trade off between the speed and the necessary details for accurate description of the relevant phenomena. The traditional implementation of PBE in molecular biology considers that biological macromolecules are low dielectric cavities immersed in water phase which is described as a continuum medium with large dielectric constant. In this protocol, Ions are modeled as point charges that obey the Boltzmann distribution and can be present outside of the so-called Stern layer (Figure \ref{fig:Figure_1}a). This implementation of PBE will be termed two-dielectric PBE. In parallel, we have implemented a Gaussian-based smooth dielectric function in DelPhi that treats the solute and solvent on the same footage and the entire computational space is described via continuous dielectric function\cite{chakravorty2018gaussian,jia2017treating,li2014modeling,li2013dielectric}. Thus, the macromolecules are considered to be inhomogeneous objects and there is no sharp boundary between solute-solvent. Ions in such a scenario are considered as point charges obeying Boltzmann distribution, but the argument of the Boltzmann function has a desolvation penalty which does not allow ions to propagate into the macromolecular interior unless there is a cavity (Figure \ref{fig:Figure_1}b). This approach will be termed Gaussian PBE. Both approaches have been used in various computational algorithms aimed at computing biologically important quantities and these resources will be described here as well along with appropriate tutorials.

The DelPhi software and associated resources are available free of charge for either download as a stand-alone code or used as web-servers. The general access URL is \url{http://compbio.clemson.edu}.

@@

\subsubsection{Continuum electrostatics quantities}
The key electrostatic quantity discussed in this tutorial is electrostatic potential distribution throughout the space. The space is occupied by macromolecule(s) and water phase. In terms of continuum electrostatics, the macromolecules and water phase are considered as continuum media. For such a system, the electrostatic potential obeys various differential equations, which will be described below. Essential term in these equations is so termed dielectric “constant” or dielectric function. In some models, the dielectric function is a constant in macromolecular and water media, while in other models it is a smooth continuous function. The macromolecules are considered at an atomistic level of details, and atomistic radii and charges are assigned according appropriate force field parameters. Finally, the mobile ions present in the water phase are considered point charges and their distribution is governed by the Boltzmann distribution. Below we outline two major models of partitioning the computational space.

\subsubsection{Space partitioning in case of traditional two-dielectric model}
In the case of traditional PBE, the space partitioning and solving corresponding differential equations involve splitting the space into three regions, labeled as $ \Omega_1 $, $ \Omega_2 $ and $ \Omega_3 $ in Figure \ref{fig:Figure_1}a. The first space region, the $\Omega_1$ is occupied by the macromolecule(s) and its boundary is defined as the molecular surface. It is considered to be a low dielectric cavity with a dielectric constant of $\epsilon_1$. Molecular surface itself may be constructed via various approaches, which are not subject of this manuscript. The outer space region, the $\Omega_3$ region, is taken by the bulk water, considered to be continuum medium with dielectric constant of $\epsilon_3$. In-between is space region $\Omega_2$, which is termed Stern layer. This region has the same dielectric properties as the bulk water, but ions are not allowed to penetrate there (Figure \ref{fig:Figure_1}a).

\begin{figure*}[hbt!]
\includegraphics[width=0.95\linewidth]{Figure_1.png}
\caption{\textbf{(a)} Schematic presentation of space partitioning in case of traditional two-dielectric model. In each of these partitions, different differential equation is solved. The domain, which is essentially the 3D space containing the macromolecule(s) and the solvent (with or without ions) is split into three regions labeled as $ \Omega_1$, $ \Omega_2 $ and $ \Omega_3 $. Colored dots are used to represent charge heterogeneity of atoms (red: negatively charged, blue: positively charged and white neutral). \textbf{(b)} Schematic presentation of the dielectric distribution delivered using the Gaussian-based smooth dielectric model (figure shows dielectric function value throughout a slice of a protein). Note that there is no dielectric boundary and even some space regions inside the protein are assigned high dielectric constant.}
\label{fig:Figure_1}
\end{figure*}

@@

The outer space region or the $ \Omega_3 $ is occupied by the solvent (typically water) which represents a continuum medium with a dielectric value $ \epsilon_3 $. The solvent continuum can also contain mobile ions whose effects are implicitly taken into account using the Boltzmann distribution. The potential distribution in this region is governed by the Poisson-Boltzmann equation (PBE). The charge density due to the mobile ions with concentration M($ \rho_3(\Vec{r}) $) generates a potential distribution that can be expressed as Eqn. \ref{eqn:pbe_1}.

@@

In the above equations, $ e_c $ is the elementary charge, $ k_B $ is the Boltzmann constant and $ T $ is the absolute temperature.
In between these two regions,  exists the ion-exclusion layer (shown as $ \Omega_2 $) which the mobile ions are not allowed to access. Due to the lack of mobile ions, as well as the absence of permanent solute charges, this region is governed by Laplace’s equation.
\begin{equation}
\nabla^2\phi(\Vec{r}) = 0
\end{equation}
The ion-exclusion layer or the Stern layer has the same dielectric properties as the solvent region or $ \Omega_3 $ (i.e. $ \epsilon_3 $). Therefore, $ \epsilon_2=\epsilon_3 $ . This indicates the role of the molecular surface as the surface that separates two dielectric media. The molecular surface can be constructed using various approaches but they are out of the scope of this article\cite{rocchia2002rapid,chakravorty2019grid,decherchi2013between}.

Briefly, Van der Waals (vdW) surface is the surface made of atomc vdW surfaces exposed to the solvent. If one runs a water probe, considered to be sphere with radius 1.4 \text{\AA}, and follow the center of the probe, the resulting surface is termed solvent accessible surface (SAS). Moving SAS 1.4 \text{\AA} towards vdW surface results in molecular surface. Finally, the ion accessible surface, which defines Stern layer, is the surface made of ions centers, represented as sphere with typical radius of 2.0 \text{\AA}.

\subsubsection{Space partitioning with the Gaussian-based smooth dielectric model}
The Gaussian-based smooth dielectric model is designed to deliver a continuous distribution of dielectric values in space as opposed to representing it as a union of two media with distinct dielectric constants. 
The smooth distribution reflects the effect of restrictions on the mobility of the macromolecular and solvent regions on their polarizability or the dielectric value. The distribution is assigned based on the macromolecular atomic packing density in a given region. Its formulation ensures that the dielectric value is the lowest ($ \epsilon_{in} $) at the center of the macromolecular atoms and is the highest ( $ \epsilon_{out} $ ) in the bulk. Through this assignment, regions close to the macromolecular atom centers attain a lower dielectric value and those farther away from the macromolecule attain a higher value. Overall, buried or highly densely packed regions of the system feature a lower dielectric value than those which are less packed (details are provided in refs\cite{chakravorty2019grid,decherchi2013between}). This is illustrated in Figure \ref{fig:Figure_1}b.

The key feature of Gaussian-based smooth dielectric functions are that the strict surface (like the molecular surface shown in Figure \ref{fig:Figure_1}a) is replaced by a smooth transition of the dielectric values from the macromolecular interior $ \Omega_1 $ to the bulk $ \Omega_3 $ as shown in Figure \ref{fig:Figure_1}. The transient values of the dielectric at the solute-solvent interface collectively reflect the increased polarizability of solvent exposed atoms of the macromolecule (compared to those in the core regions) and decreased polarizability of the solvent in their vicinity (compared to those in the bulk). This occurs by virtue of the solute-solvent interactions that play a critical role in balancing the effects of (de)solvation and intramolecular interactions of the solvent exposed macromolecular atoms. 
Due to the continuous dielectric distribution, a clear distinction of macromolecular and solvent regions is no longer necessary and therefore, the same differential equation (PE or PBE) can be used homogeneously. However, the lack of a dielectric boundary invites a modification in the way by which the screening effect of mobile ions is taken into account. In the traditional two-dielectric approach, the presence of a mobile ion at a point in space (depending on its accessibility) is guided by the electrostatic potential at that point which is expressed using a Boltzmann distribution function (Eqn. \ref{eqn:pbe_2}). In the Gaussian-based approach, an additional penalty term is added to its potential energy at any point in space (for more details see Ref\cite{jia2017treating}). The penalty term is based on the Born-formalism of the energy of transfer of a centrosymmetric ion across two dielectric media. In the case of the Gaussian-model, the penalty term, $ \Delta G_{penalty} $, is the energy of transfer of an ion of valence $ z $ and Van der Waals radius $ R $ from the solvent (with dielectric $ \epsilon_2 $) to a point  (with dielectric $ \epsilon(\Vec{r})$) and is expressed in SI units as:

@@

In Figure \ref{fig:plot_saltation_penalty}, the penalty term as a function of position is shown for a solute with a very simple geometry (one sphere of 2 \text{\AA} radius). As one approaches the vicinity of the solute, the penalty increases dramatically, ensuring that the mobile ions cannot penetrate in the regions dominated by solute’s dielectric values. Farther out in the bulk, the penalty term diminishes to zero, thus retrieving the properties of the traditional 2-dielectric approach.

@@

\subsubsection{Finite difference algorithm}
The Poisson-Boltzmann equation is a non-linear $2^{nd}$ order differential equation that is typically solved using numerical schemes since analytical solutions are only possible for cases with very simple geometry. Inhomogeneity of charge distribution and complex geometrical features are prerogative of biomolecules, therefore, different methods of numerically solving the PBE for biomolecular systems have been developed \cite{klapper1986focusing,davis1989solving,cortis1997numerical,baker2001electrostatics,shestakov2002solution,miertuvs1981electrostatic,totrov2001rapid,lu2009adaptive}. 
DelPhi solves the PBE using the finite difference (FD) algorithm. The approach rests on discretization of the entire space, $ \Omega=(\Omega_1 \cup \Omega_2 \cup \Omega_3)$, into a set of adjacent cubes, all of the same side length. Along each of the three Cartesian orthogonal directions, an identical number of these cubes are placed and their union forms a larger cube which is also referred to as the computational box. Points where two or more cubes meet are known as the grid points and they are the locations where the electrostatic potentials and electric fields are determined.  Between two adjacently placed grid-points, a mid-point is located where the dielectric values are assigned. Collectively, the grid-points and mid-points are all placed within the bounds of the computational box. Figure \ref{fig:grid-scheme-representation}, provides an illustration of this discretization and a computational box with ‘$ M $’ cubes per side, each with a side length of ‘$ h $’.

@@

Instead of solving different variants of the differential equation like the PE and PBE (as mentioned above) in the different partitions or subdomains of the computational box, an iterative protocol based on the integral form of these equations is executed to compute the potential values at each grid-point. To do so, a position dependent formulation of the dielectric values ($ \epsilon $) and the modified Debye-Hückel parameter ($ \kappa $) is used.

@@

\subsection{Energy partitioning}
Once the potential distribution is calculated, the next step in DelPhi algorithm is to compute the electrostatic energies. Based on the original paper of Sharp and Honig\cite{sharp1990electrostatic}, see also\cite{rocchia2001extending}, the energy components are:
Grid energy
\begin{equation}
    G_{grid} = 0.5 \left[ \sum_i^{grids}{p_i\phi_i} \right]
\end{equation}
where $ p_i $ is the charge of grid point `$ i $' and $ \phi_i $ is the grid potential at the same grid point. Note that this energy term contains artificial components due to the real charge partitioning over the grid points. Direct usage of $ G_{grid} $ is not appropriate, only grid energy difference can be used.

The total electrostatic energy of the system can be decomposed into three energy terms\cite{rocchia2001extending}, Coulombic, polar solvation and saltation energies. In case of traditional two-dielectric model, this is described in Ref. \cite{rocchia2001extending}. Thus, the Coulombic term is the product between real charges over the distance, and it is computed via Coulombic formula:

\begin{equation}
    G_{Coul} = 0.5 \left[ \sum_i^{\text{number of charges}}{\frac{q_i q_j}{\epsilon_{in}r_{ij}}} \right]
\end{equation}	
where $ q_i $ are the real charges, $r_{ij} $ is the distance between charge `$i$' and `$j$', and $ \epsilon_{in} $ is the internal dielectric constant.

The second energy term is the polar solvation energy, which in DelPhi is called ``corrected reaction field energy'', or ``$ G_{rxn} $''. Following original development due to Anthony Nicholls, this energy is calculated via induced charges on the molecular surface. The induced charges ($ \sigma_j $) are calculated using electrostatic potential distribution over the molecular surface. The $ \sigma_j $ are obtained via Gauss's Law at the corresponding grid point and then projected to the molecular surface (more details are provided in \cite{rocchia2001extending}). Thus,

@@

The third term represents the energy of interactions between permanent charges and mobile ions. Formally it could be described by a formula similar to Eqn. \ref{eqn:G_rxn}, where instead of induced charges one takes excess ion density. However, this approach was abandoned in the newest DelPhi version because it is very slow and requires small “percent fill”. Instead, we recommend using grid energy difference between runs with salt and without salt.

Finally, if one invokes non-linear PBE, two additional energy terms should be added to the total electrostatic energy. Following the work of Sharp and Honig\cite{sharp1990electrostatic}, they are termed ``osmotic pressure'' and ``electrostatic stress''.

@@

DelPhi uses FD algorithm to solve the PBE and thus to deliver the electrostatic potential distribution. The first step in the algorithm requires that one provides a coordinate file of the biomolecules and nano-objects in PDB or PQR formats. In the second case, the input PQR file provides the coordinates of all atoms along with their radii and charges. In the first case, the user needs to provide two additional files, the ``size'' and ``charge'' files. These files contain information about the atomic radii and charges. If PDB format file is provided, then DelPhi matches the names of the atoms/residues toward the names/radii in the corresponding ``size'' and ``charge'' files and thus generates information for each atom position, radius and charge (note that if the PDB file contains atomic names/residues not available in the corresponding ``size'' and ``charge'' files, then DelPhi will generate warning message prompting the users that these names/residues are not recognized). Once the atomic coordinates and radii are available, DelPhi builds grid presentation of the geometrical and dielectric properties of the system. By default, it is done by positioning the macromolecule(s) geometrical center at the center of the computational box and using ``scale'', ``grid size'' and ``percent fill''.

\begin{figure}[hbt!]
\includegraphics[width=\linewidth]{Figure_4.png}
\caption{The schematic representation of relationship between ``percent fill'' (\ttvar{perfil}), ``grid size'' (\ttvar{gsize}), and \ttvar{scale}. The parameter \ttvar{gsize} decides number of grid lines in each direction e.g. \ttvar{gsize} is 9 in given cartoon. The parameter \ttvar{scale} decides number of grid lines per \text{\AA}. If we assume \ttvar{perfil} $ pf\% $ and maximum dimension in X-, Y- directions be $ L_{xp} $ \text{\AA} and $ L_{yp} $ \text{\AA} as shown in cartoon, then the size of the grid box $ L_x $ in \text{\AA} is computed using relation $ L_x = \frac{max(L_{xp}, L_{yp}) \times 100}{pf} $. The value of \ttvar{scale} in this case would be $ \ttvar{scale} = \frac{L_x}{\ttvar{gsize}} = \frac{L_x}{9} $.}
\label{fig:perfil-gsize-scale}
\end{figure}

Note that these three parameters, ``scale'', ``grid size'' and ``percent fill'' are interdependent (Figure \ref{fig:perfil-gsize-scale}). Only two are needed. Users are advised to use ``scale'' and ``percent fill'' (\ttvar{perfil}) and to let DelPhi decide the ``grid size'' (\ttvar{gsize}). In the traditional two-dielectric model, the computational box is split into low dielectric cavities occupied by macromolecule(s) and water phase and appropriate dielectric constant values are assigned to the mid-point belonging to the corresponding phases. In the Gaussian model, coordinates and radii are used to construct a smooth Gaussian-based dielectric function for the entire computational box without distinguishing the macromolecule(s) from the water phase (for mode details see Refs \cite{li2014modeling,li2013dielectric}). Once this is completed, DelPhi initiates FD iterations to compute electrostatic potential distribution and to calculate the corresponding energy, as it will be outlined below. Several practical advice about DelPhi parameters are given here:

\begin{enumerate}
    \item How to decide on geometry? One wants the macromolecule(s) to occupy significant percent of the computational box space so the calculations to focus on important part of the system. However, doing modeling with very high ``percent fill'' may result in positioning some of the macromolecule(s) atoms too close to the edges of the computational box and thus to introduce an error. We recommend ``percent fill'' of about 70\%.

    \item How to decide on the ``dielectric constants''? Most of the modeling involves biological macromolecules in water, so natural choice of the dielectric constant of solvent is the dielectric constant of bulk water, typically taken as 80. Regarding internal dielectric constant: (a) in case of traditional two-dielectric model, one selects the value of internal dielectric constant depending on the modeling protocol. If one models rigid structures or snap shots taken from MD simulations, the internal dielectric constant value is typically set to 1.0. However, in other modelings one may select values that are higher with the goal to match experimental data or to mimic for conformational flexibility not accounted in the protocol (b) in case of smooth Gaussian-based dielectric function, one selects the reference value of internal dielectric constant based on benchmarking (see refs\cite{li2014modeling,li2013progress,wang2015pka,wang2015delphipka} for more details).

    \item Criteria for convergence. The FD algorithm is iterative and its convergence is monitored via several parameters as ``maxc'', which is the largest change of the potential from iteration `i' to `i+1'; or ``rmsc'', which root mean square change of the potential. In both cases, it is recommended to use small, but reasonable criteria of 0.0001.

@@

Now close your terminal and restart the terminal. Now the user shall be able to perform calculations discussed in all of the examples compiled in directory \ttvar{LiveCoMS_Examples/}. All the paths mentioned in individual sections are relative to this directory.
